---
title: "Location of MPRA variants in repeats, Vijay's data"
author: "Grace Hansen"
date: "06/29/2019"
output: pdf_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/midway/MPRA_repeats/')
library(data.table)
library(tidyverse)
library(gridExtra)
pony_colors<-fread("~/medusa/papers/TWAS/pony_palette")
########## For color manipulation of graph ###############
darken <- function(color, factor=1.2){
  col <- col2rgb(color)
  col <- col/factor
  col <- rgb(t(col), maxColorValue=255)
  col
}
lighten <- function(color, factor=0.8){
  col <- col2rgb(color)
  col <- col/factor
  for (i in 1:length(col)) { if ( col[i] > 255 ) { col[i] = 255 } }
  col <- rgb(t(col), maxColorValue=255)
  col
}
#########################################################
```


```{r load_data,echo=FALSE}
EMVar_repeats=system("bedtools intersect -a rmsk_hg38.bed -b ~/projects/MPRA/vijay/vijay_half_EMVars_hg38.bed -wa -wb",intern=TRUE)
EMVar_repeats<-data.frame(do.call('rbind',strsplit(as.data.frame(EMVar_repeats,stringsAsFactors=FALSE)$EMVar_repeats,'\t',fixed=TRUE)))
colnames(EMVar_repeats)<-c('chr','rep_start','rep_stop','rep_type','rep_score','rep_strand','rep_subtype','chr1','pos','pos1')
EMVar_repeats=EMVar_repeats[,c('rep_start','rep_stop','rep_type','rep_score','rep_strand','rep_subtype','chr','pos')]
EMVar_repeats$pos<-as.numeric(as.character(EMVar_repeats$pos))
EMVar_rsids<-fread("~/projects/MPRA/vijay/vijay_half_EMVars_hg38.bed",header=FALSE)
colnames(EMVar_rsids)<-c('chr','pos','pos1')
EMVar_repeats=merge(EMVar_rsids[,c('chr','pos')],EMVar_repeats,by=c("chr","pos"),all.x=TRUE)
EMVar_repeats$type="EMVar"

enhancer_repeats=system("bedtools intersect -a rmsk_hg38.bed -b ~/projects/MPRA/vijay/vijay_half_enhancers_hg38.bed -wa -wb",intern=TRUE)
enhancer_repeats<-data.frame(do.call('rbind',strsplit(as.data.frame(enhancer_repeats,stringsAsFactors=FALSE)$enhancer_repeats,'\t',fixed=TRUE)))
colnames(enhancer_repeats)<-c('chr','rep_start','rep_stop','rep_type','rep_score','rep_strand','rep_subtype','chr1','pos')
enhancer_repeats=enhancer_repeats[,c('rep_start','rep_stop','rep_type','rep_score','rep_strand','rep_subtype','chr','pos')]
enhancer_repeats$pos<-as.numeric(as.character(enhancer_repeats$pos))
enhancer_rsids<-fread("~/projects/MPRA/vijay/vijay_half_enhancers_hg38.bed",header=FALSE)
colnames(enhancer_rsids)<-c('chr','pos','pos1')
enhancer_repeats=merge(enhancer_rsids[,c('chr','pos')],enhancer_repeats,by=c("chr","pos"),all.x=TRUE)
enhancer_repeats$type="enhancer"

nonsig_repeats=system("bedtools intersect -a rmsk_hg38.bed -b ~/projects/MPRA/vijay/vijay_half_nonsig_hg38.bed -wa -wb",intern=TRUE)
nonsig_repeats<-data.frame(do.call('rbind',strsplit(as.data.frame(nonsig_repeats,stringsAsFactors=FALSE)$nonsig_repeats,'\t',fixed=TRUE)))
colnames(nonsig_repeats)<-c('chr','rep_start','rep_stop','rep_type','rep_score','rep_strand','rep_subtype','chr1','pos','pos1')
nonsig_repeats=nonsig_repeats[,c('rep_start','rep_stop','rep_type','rep_score','rep_strand','rep_subtype','chr','pos')]
nonsig_repeats$pos<-as.numeric(as.character(nonsig_repeats$pos))
nonsig_rsids<-fread("~/projects/MPRA/vijay/vijay_half_nonsig_hg38.bed",header=FALSE)
colnames(nonsig_rsids)<-c('chr','pos','pos1')
nonsig_repeats=merge(nonsig_rsids[,c('chr','pos')],nonsig_repeats,by=c("chr","pos"),all.x=TRUE)
nonsig_repeats$type="nonsig"

dat=rbind(EMVar_repeats,enhancer_repeats,nonsig_repeats)
EMVar_pos<-dat$pos[dat$type=="EMVar"]
dat<-dat[!(dat$type=="enhancer" & dat$pos %in% EMVar_pos),]
dat<-dat %>% distinct(chr,pos,.keep_all=TRUE) #Picking only unique matches (throws out a few entries where there are multiple repeat subtypes)
write.table(dat,"~/midway/MPRA_repeats/MPRA_variants_repeats_vijay.txt",row.names=FALSE,sep='\t',quote=FALSE)

```

### Location of MPRA variants in repeats
```{r fraction_repeats,echo=FALSE}
summary=matrix(nrow=3,ncol=4)
j=1
for (i in c("EMVar","enhancer","nonsig")) {
  n_vars=length(dat$rep_type[dat$type==i])
  prop_rep=round(sum(!(is.na(dat$rep_type[dat$type==i])))/length(dat$rep_type[dat$type==i]),digits=4)
  prop_alu=round(nrow(dat[dat$type==i & grepl("Alu",dat$rep_subtype),])/length(dat$rep_type[dat$type==i]),digits=4)
  summary[j,]<-c(i,n_vars,prop_rep,prop_alu)
  j=j+1
}
summary=as.data.frame(summary)
colnames(summary)<-c("Variant type","Number of variants","Proportion in repeats","Proportion in Alus")
summary$`Proportion in repeats`<-as.numeric(as.character(summary$`Proportion in repeats`))
summary$`Proportion in Alus`<-as.numeric(as.character(summary$`Proportion in Alus`))
summary$`Variant type`<-as.character(summary$`Variant type`)
summary$`Variant type`[summary$`Variant type`=="enhancer"]<-"Enhancer"
summary$`Variant type`[summary$`Variant type`=="nonsig"]<-"Nonsignificant"
knitr::kable(summary)
```

### Retrovirus types in MPRA
Types of repeats in EMVars:
```{r repeat_types,echo=FALSE}
knitr::kable(sort(table(dat$rep_type[dat$type=="EMVar"]),decreasing=TRUE),col.names=c('RV','count'))
```

Sub-types of repeats in EMVars:
```{r repeat_subtypes,echo=FALSE}
knitr::kable(head(sort(table(dat$rep_subtype[dat$type=="EMVar" & grepl("Alu",dat$rep_subtype)]),decreasing=TRUE),27),col.names=c('RV subtype','count'))
```

### Plot results with chi-square p values
Are there more repeats and Alus in repeats than expected by chance?

```{r plot_results,echo=FALSE}
Alu_prop<-nrow(dat[grepl("Alu",dat$rep_subtype),])/nrow(dat)
repeat_prop<-nrow(dat[dat$rep_type!=0,])/nrow(dat)
dat <- data.frame(lapply(dat, as.character), stringsAsFactors=FALSE)
dat[is.na(dat)]<-0

# Repeats
expected<-matrix(nrow=3,ncol=2)
colnames(expected)<-c("repeats","no repeats")
rownames(expected)<-c("EMVars","enhancers","nonsig")
observed<-expected
expected[1,1]<-nrow(dat[dat$type=="EMVar",])*repeat_prop
expected[1,2]<-nrow(dat[dat$type=="EMVar",])*(1-repeat_prop)
expected[2,1]<-nrow(dat[dat$type=="enhancer",])*repeat_prop
expected[2,2]<-nrow(dat[dat$type=="enhancer",])*(1-repeat_prop)
expected[3,1]<-nrow(dat[dat$type=="nonsig",])*repeat_prop
expected[3,2]<-nrow(dat[dat$type=="nonsig",])*(1-repeat_prop)

observed[1,1]<-nrow(dat[dat$type=="EMVar" & dat$rep_type!=0,])
observed[1,2]<-nrow(dat[dat$type=="EMVar" & dat$rep_type==0,])
observed[2,1]<-nrow(dat[dat$type=="enhancer" & dat$rep_type!=0,])
observed[2,2]<-nrow(dat[dat$type=="enhancer" & dat$rep_type==0,])
observed[3,1]<-nrow(dat[dat$type=="nonsig" & dat$rep_type!=0,])
observed[3,2]<-nrow(dat[dat$type=="nonsig" & dat$rep_type==0,])

print("Repeat Chi square observed vs expected tables:")
observed
expected

repeat_p<-chisq.test(observed,expected)$p.value

#Alus
expected<-matrix(nrow=3,ncol=2)
colnames(expected)<-c("repeats","no repeats")
rownames(expected)<-c("EMVars","enhancers","nonsig")
observed<-expected
expected[1,1]<-nrow(dat[dat$type=="EMVar",])*Alu_prop
expected[1,2]<-nrow(dat[dat$type=="EMVar",])*(1-Alu_prop)
expected[2,1]<-nrow(dat[dat$type=="enhancer",])*Alu_prop
expected[2,2]<-nrow(dat[dat$type=="enhancer",])*(1-Alu_prop)
expected[3,1]<-nrow(dat[dat$type=="nonsig",])*Alu_prop
expected[3,2]<-nrow(dat[dat$type=="nonsig",])*(1-Alu_prop)

observed[1,1]<-nrow(dat[dat$type=="EMVar" & grepl("Alu",dat$rep_subtype),])
observed[1,2]<-nrow(dat[dat$type=="EMVar" & !(grepl("Alu",dat$rep_subtype)),])
observed[2,1]<-nrow(dat[dat$type=="enhancer" & grepl("Alu",dat$rep_subtype),])
observed[2,2]<-nrow(dat[dat$type=="enhancer" & !(grepl("Alu",dat$rep_subtype)),])
observed[3,1]<-nrow(dat[dat$type=="nonsig" & grepl("Alu",dat$rep_subtype),])
observed[3,2]<-nrow(dat[dat$type=="nonsig" & !(grepl("Alu",dat$rep_subtype)),])

print("Alu Chi square observed vs expected tables:")
observed
expected

Alu_p<-chisq.test(observed,expected)$p.value

## Get genome proportions
repeat_genome_prop<-scan("~/midway/MPRA_repeats/repeat_prop")
Alu_genome_prop<-scan("~/midway/MPRA_repeats/Alu_prop")
summary<-as.data.frame(rbind(summary,c("Genome proportion",NA,repeat_genome_prop,Alu_genome_prop)),stringsAsFactors=FALSE)
summary$`Proportion in repeats`<-as.numeric(summary$`Proportion in repeats`)
summary$`Proportion in Alus`<-as.numeric(summary$`Proportion in Alus`)
summary$`Variant type`<-factor(summary$`Variant type`,levels=c("EMVar","Enhancer","Nonsignificant","Genome proportion"))

## Plot results
pdf("MPRA_vars_in_repeats_vijay.pdf",width=7,height=3)
ggplot(summary)+
  geom_col(aes(x=`Variant type`,y=`Proportion in repeats`,fill=`Variant type`))+
  annotate("text",x=2.65,y=0.85,label=paste("Chisq p= \n",formatC(repeat_p,digits=4),sep=''),cex=6)+
  theme_minimal()+
  scale_y_continuous(limits=c(0,1))+
  scale_x_discrete(name='')+
  scale_fill_manual(values=c(darken(rgb(pony_colors[3,1:3])),darken(rgb(pony_colors[5,1:3])),darken(rgb(pony_colors[9,1:3])),"gray80"))+
  theme(axis.text.x=element_text(size=20,angle=-45),
        axis.text.y=element_text(size=20),
        axis.title=element_text(size=20),
        legend.position="none")+
  coord_flip()
dev.off()

pdf("MPRA_vars_in_Alus_vijay.pdf",width=7,height=3)
ggplot(summary)+
  geom_col(aes(x=`Variant type`,y=`Proportion in Alus`,fill=`Variant type`))+
  annotate("text",x=2.65,y=0.8,label=paste("Chisq p= \n",formatC(Alu_p,digits=4),sep=''),cex=6)+
  theme_minimal()+
  scale_y_continuous(limits=c(0,1))+
  scale_x_discrete(name='')+
  scale_fill_manual(values=c(darken(rgb(pony_colors[3,1:3])),darken(rgb(pony_colors[5,1:3])),darken(rgb(pony_colors[9,1:3])),"gray80"))+
  theme(axis.text.x=element_text(size=20,angle=-45),
        axis.text.y=element_text(size=20),
        axis.title=element_text(size=20),
        legend.position="none")+
  coord_flip()
dev.off()

```

## Expression of Alu elements

Are sequences containing Alu elements more highly expressed than sequences not containing Alu elements?

```{r summary_of_distribution,echo=FALSE}
dat$type[dat$type=="nonsig" & !(grepl("Alu",dat$rep_subtype))]<-"Nonsig, no Alus"
dat$type[dat$type=="nonsig" & grepl("Alu",dat$rep_subtype)]<-"Nonsig, Alus" #SINE has to be second: some EMVars are in two elements if I use -F 0.25
dat$type[dat$type=="enhancer" & !(grepl("Alu",dat$rep_subtype))]<-"Enhancer, no Alus"
dat$type[dat$type=="enhancer" & grepl("Alu",dat$rep_subtype)]<-"Enhancer, Alus"
dat$type[dat$type=="EMVar" & !(grepl("Alu",dat$rep_subtype))]<-"EMVar, no Alus"
dat$type[dat$type=="EMVar" & grepl("Alu",dat$rep_subtype)]<-"EMVar, Alus"

qnorm<-as.data.frame(fread("~/projects/MPRA/vijay/MPRA_qnorm.txt"),stringsAsFactors=FALSE)
qnorm<-qnorm[qnorm$bot=="1/2",]
qnorm$qnorm_avg<-rowMeans(qnorm[,grepl("RATIO",colnames(qnorm))])
qnorm<-qnorm %>% group_by(chr,pos,ref,alt,oligo,construct) %>% summarize(qnorm_avg=mean(qnorm_avg))
qnorm$chr=paste("chr",qnorm$chr,sep='')
qnorm$pos1=qnorm$pos+1
write.table(qnorm[,c("chr","pos","pos1","qnorm_avg")],"~/projects/MPRA/vijay/MPRA_qnorm_avg_hg19.bed",quote=FALSE,sep='\t',row.names=FALSE,col.names=FALSE)
system("liftOver ~/projects/MPRA/vijay/MPRA_qnorm_avg_hg19.bed ~/medusa/reference/hg19ToHg38.over.chain.gz ~/projects/MPRA/vijay/MPRA_qnorm_avg_hg38.bed unMapped")
qnorm<-as.data.frame(fread("~/projects/MPRA/vijay/MPRA_qnorm_avg_hg38.bed"),stringsAsFactors=FALSE)
colnames(qnorm)<-c("chr","pos","pos1","qnorm_avg")
dat<-merge(dat,qnorm[,c("chr","pos","qnorm_avg")],by=c("chr","pos"))

print("qnorm from nonsignificant sequences without Alu elements")
summary(dat$qnorm_avg[dat$type=="Nonsig, no Alus"])
print("qnorm from nonsignificant sequences with Alu elements")
summary(dat$qnorm_avg[dat$type=="Nonsig, Alus"])
print("qnorm from enhancer sequences without Alu elements")
summary(dat$qnorm_avg[dat$type=="Enhancer, no Alus"])
print("qnorm from enhancer sequences with Alu elements")
summary(dat$qnorm_avg[dat$type=="Enhancer, Alus"])
print("qnorm from EMVar sequences without Alu elements")
summary(dat$qnorm_avg[dat$type=="EMVar, no Alus"])
print("qnorm from EMVar sequences with Alu elements")
summary(dat$qnorm_avg[dat$type=="EMVar, Alus"])
```

### Significance and visualization

In the plots below, you can see that the Alu-containing sequences have higher expression in enhancers and EMVars, but this isn't true for nonsignificant sequences.


```{r plot_alus,echo=FALSE}

nonsig_ttest<-t.test(dat$qnorm_avg[dat$type=="Nonsig, Alus"],dat$qnorm_avg[dat$type=="Nonsig, no Alus"])
enhancer_ttest<-t.test(dat$qnorm_avg[dat$type=="Enhancer, Alus"],dat$qnorm_avg[dat$type=="Enhancer, no Alus"])
EMVar_ttest<-t.test(dat$qnorm_avg[dat$type=="EMVar, Alus"],dat$qnorm_avg[dat$type=="EMVar, no Alus"])

M<-ggplot(dat[grepl("EMVar",dat$type),])+
  geom_violin(aes(x=type,y=qnorm_avg,fill=type),draw_quantiles=c(0.5))+
  scale_x_discrete(name='')+
  scale_y_continuous(name="Average activity per variant")+
  annotate("text",x=1.18,y=8.75,label=paste("p=",formatC(EMVar_ttest$p.value, format = "e", digits = 3),sep=''),cex=6)+
  scale_fill_manual(values=c(darken(rgb(pony_colors[3,1:3])),lighten(rgb(pony_colors[3,1:3]))))+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=-45),
        axis.text=element_text(size=16),
        axis.title=element_text(size=15),
        legend.position = "none")
E<-ggplot(dat[grepl("Enhancer",dat$type),])+
  geom_violin(aes(x=type,y=qnorm_avg,fill=type),draw_quantiles=c(0.5))+
  scale_x_discrete(name='')+
  scale_y_continuous(name="Average activity per variant")+
  annotate("text",x=1.18,y=9.25,label=paste("p=",formatC(enhancer_ttest$p.value, format = "e", digits = 3),sep=''),cex=6)+
  scale_fill_manual(values=c(darken(rgb(pony_colors[5,1:3])),lighten(rgb(pony_colors[5,1:3]))))+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=-45),
        axis.text=element_text(size=16),
        axis.title=element_text(size=15),
        legend.position = "none")
N<-ggplot(dat[grepl("Nonsig",dat$type),])+
  geom_violin(aes(x=type,y=qnorm_avg,fill=type),draw_quantiles=c(0.5))+
  scale_x_discrete(name='')+
  scale_y_continuous(name="Average activity per variant")+
  annotate("text",x=1.18,y=3.5,label=paste("p=",formatC(nonsig_ttest$p.value, format="e",digits = 3),sep=''),cex=6)+
  scale_fill_manual(values=c(darken(rgb(pony_colors[9,1:3])),lighten(rgb(pony_colors[9,1:3]))))+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=-45),
        axis.text=element_text(size=16),
        axis.title=element_text(size=15),
        legend.position = "none")

grid.arrange(M,E,N,nrow=1)

pdf("Alus_vs_noAlus_vijay.pdf",width=8,height=5)
grid.arrange(M,E,N,nrow=1)
dev.off()


```
